Copy number inference with InferCNV

### 2024-05-13 analyses/cell-type-ewings/01-marker-gene-tumor-classification.Rmd
move files

```{r}
# source in helper functions: make_jaccard_matrix and jaccard 
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

@@

# pull out the UMAP coordinates and make a data frame to use for plotting
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP"),
        starts_with("UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, cluster)
```

>>>>

```{r}
# source in helper functions: make_jaccard_matrix() and jaccard() 
jaccard_functions <- file.path(module_base, "scripts", "utils", "jaccard-functions.R")
source(jaccard_functions)
```

@@

# pull out the UMAP coordinates and make a data frame to use for plotting
umap_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      ) |> 
  # add in marker gene expression to dataframe 
  dplyr::left_join(marker_gene_exp, by = "barcodes") |> 
  # combine all genes into a single column for easy faceting
  tidyr::pivot_longer(
    cols = starts_with("ENSG"),
    names_to = "ensembl_gene_id",
    values_to = "gene_expression"
  ) |> 
  # join with marker gene df to get gene symbols for plotting 
  dplyr::left_join(marker_genes_df, by = c("ensembl_gene_id")) |>
  dplyr::select(barcodes, UMAP1, UMAP2, gene_symbol, ensembl_gene_id, gene_expression, cluster)
```

### 2024-05-13 analyses/cell-type-ewings/03-copykat.Rmd
move files

`CopyKAT` outputs a matrix that contains the estimated copy numbers for each gene in each cell. 
We can read that in and look at the mean estimated copy numbers for each chromosome across each cell. 
We might expect that tumor cells would show an increased estimated copy number in Chr8, Chr12, and/or Chr 1 and a loss of Chr16.

@@

# join with cnv info
cnv_df <- cnv_df |> 
  dplyr::left_join(full_cnv_df, by = c("barcodes", "reference_used"))
```

Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.

```{r, fig.height=15, fig.width=10}
# create faceted density plots showing estimation of CNV detection across each chr of interest 
# colored by aneuploid/diploid estimation 
ggplot(cnv_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
      geom_density() +
      theme_bw() +
      facet_grid(rows = vars(chromosome_name), 
                 cols = vars(reference_used))
```

It looks like the largest increases are in chr8 and chr21. 
Chr8 is one of the more common CNA's in Ewing's so this is a good sign.

Chr12 may show some increase, but Chr1 and Chr16 don't show any real differences.

>>>>

`CopyKAT` outputs a matrix that contains the estimated copy numbers for each gene in each cell. 
We can read that in and look at the mean estimated copy numbers for each chromosome across each cell. 
We might expect that tumor cells would show an increased estimated copy number in Chr8, Chr12, and/or Chf 1 and a loss of Chr16.

@@

# filter to only have cnvs found in Ewings to keep things a bit smaller when joining
ewings_chrs <- c(1, 8, 12, 16)

ewings_cnv_df <- full_cnv_df |> 
  dplyr::filter(chromosome_name %in% ewings_chrs)

# join with 
cnv_df <- cnv_df |> 
  dplyr::left_join(ewings_cnv_df, by = c("barcodes", "reference_used"))
```

```{r}
# create faceted UMAPs showing estimation of CNV detection across each chr of interest 
ewings_chrs |> 
  purrr::map(\(chr){
    plot_df <- cnv_df |> 
      dplyr::filter(chromosome_name == chr)

    ggplot(plot_df, aes(x = UMAP1, y = UMAP2, color = mean_cnv_detection)) +
      geom_point(alpha = 0.5, size = 0.5) +
      theme_bw() +
      scale_color_viridis_c() +
      facet_wrap(vars(reference_used)) + 
      labs(title = glue::glue("chr{chr}"))
  })
```

Looking at the UMAPs, I don't see any really obvious differences between CNV estimation and groups of cells. 
Let's look at the distribution of CNV estimation in cells that are called aneuploid and diploid by `CopyKAT`.

```{r}
# create faceted density plots showing estimation of CNV detection across each chr of interest 
# colored by aneuploid/diploid estimation 
ewings_chrs |> 
  purrr::map(\(chr){
    plot_df <- cnv_df |> 
      dplyr::filter(chromosome_name == chr)

    ggplot(plot_df, aes(x = mean_cnv_detection, color = copykat.pred)) +
      geom_density() +
      theme_bw() +
      facet_wrap(vars(reference_used)) + 
      labs(title = glue::glue("chr{chr}"))
  })
```

It looks like there is a slight increase in chr8 and chr12 indicating maybe a copy number gain for those chromosomes in aneuploid cells, but nothing striking.

### 2024-05-13 analyses/cell-type-ewings/04-infercnv.Rmd
move files

# output from running infercnv 
infer_cnv_dir <- file.path(module_base, "results", "infercnv", "SCPCL000822")

# add code to read in both ref and no ref? 
```

@@

Below is the heatmap produced by `InferCNV` showing the CNV's for each cell. 
Each row is a cell and each column is a gene. 
The top group contains all cells that were used as the reference group and the bottom is all other cells. 
The color indicates the relative expression intensity of each gene with the darker colors either indicating a gain of expression or loss of expression compared to the reference cells.

![InferCNV Heatmap - Endothelial cells as reference](results/infercnv/SCPCL000822/infercnv.png)
From this heatmap we see that there is a distinct group of cells within the bottom portion of the plot that appear to have a lot more CNVs than the rest of the cells. 
It's probable that these correspond to the tumor cells, but `InferCNV` does not provide any classification of cells as tumor or normal, it just provides information on copy number variations.

@@

```{r}
# first create a Seurat object from our SCE object
seurat_obj <- scpcaTools::sce_to_seurat(sce)

@@

```{r}
coldata_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      )

# remove MT and GL chr 
chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
cnv_df <- coldata_df[, chr_columns_to_keep]
```

@@

@@

@@

@@

@@

@@

@@

@@

Now when we look at the proportion, we see that only the group of tumor cells on the bottom seem to have a higher proportion of genes in chr8 that have a gain in expression. 
Based on the papers, we should see a gain in the whole chr8 and chr12, not just one arm or specific genes, so we expect a high value for the proportion. 
We see this for chr8 but not in chr12.

I think with the combination of marker gene expression and CNVs, the bottom group of cells represent the tumor cells.

## Do the same thing but without a reference set of cells?

@@

>>>>

# output from running infercnv 
results_dir <- file.path(module_base, "results", "infercnv", params$library_id)

all_results_dir_list <- c(
  no_ref = file.path(results_dir, "no_reference"),
  with_ref = file.path(results_dir, "endothelial_reference")
)
```

@@

### Endothelial cells as reference

Below is the heatmap produced by `InferCNV` showing the CNV's for each cell using endothelial cells as a reference. 
Each row is a cell and each column is a gene. 
The top group contains all cells that were used as the reference group and the bottom is all other cells. 
The color indicates the relative expression intensity of each gene with the darker colors either indicating a gain of expression or loss of expression compared to the reference cells.

![InferCNV Heatmap - Endothelial cells as reference](`r file.path(all_results_dir_list["with_ref"], "infercnv.png")`)

From this heatmap we see that there is a distinct group of cells within the bottom portion of the plot that appear to have a lot more CNVs than the rest of the cells. 
It's probable that these correspond to the tumor cells, but `InferCNV` does not provide any classification of cells as tumor or normal, it just provides information on copy number variations.

### No reference

Below is the heatmap produced by `InferCNV` showing the CNV's for each cell without using a reference.

From the [`InferCNV` documentation](https://github.com/broadinstitute/inferCNV/wiki/Running-InferCNV):

> Note, if you do not have reference cells, you can set ref_group_names=NULL, in which case the average signal across all cells will be used to define the baseline. This can work well when there are sufficient differences among the cells included (ie. they do not all show a chromosomal deletion at the same place).

Given this information, I wouldn't expect the CNV inference to work well because we see that all cells that do have prominent CNVs have the same CNVs. 
There is not a lot of variation.

![InferCNV Heatmap - No reference](`r file.path(all_results_dir_list["no_ref"], "infercnv.png")`)

Looking at this we see a small group of cells that may have some variations, but they are different from the variations seen above. 
I don't know that using this without a reference is the best call since it's unlikely that it will identify distinct groups of cells from tumors with quiet genomes with no baseline. 
The rest of the analysis will use the results from running `InferCNV` with endothelial cells as the reference.

@@

```{r}
infer_cnv_dir <- all_results_dir_list["with_ref"]

# first create a Seurat object from our SCE object
seurat_obj <- scpcaTools::sce_to_seurat(sce)

@@

@@

```{r}
# make df for plotting
coldata_df <- sce |> 
  scuttle::makePerCellDF(use.dimred = "UMAP") |> 
  # replace UMAP.1 with UMAP1
  dplyr::rename_with(
        \(x) stringr::str_replace(x, "^UMAP\\.", "UMAP")
      )

# remove MT and GL chr, comments in docs recommend that these are not useful and can be inaccurate 
chr_columns_to_keep <- which(!stringr::str_detect(names(coldata_df), "chrMT|chrGL000219.1"))
cnv_df <- coldata_df[, chr_columns_to_keep]
```

@@

For the rest of the notebook, we will use the classifications of tumor cells based on the sum of the total CNV occurrences.

@@

Below we can compare these annotations from InferCNV to classifying tumor cells manually with marker genes.

@@

This looks pretty consistent with what we see when we annotate cells manually.

We can also compare to the existing annotations obtained using `SingleR` and `CellAssign` as part of `scpca-nf`.

@@

This looks as we expect with mostly muscle cells lining up with tumor cells.

@@

Below we will plot whether or not the gain or loss is present in each cell according to `InferCNV`.

@@

We can also look at the proportion of the chromosome with the gain.

@@

Now when we look at the proportion, we see that only the group of tumor cells on the bottom seem to have a higher proportion of genes in chr8 that have a gain in expression. 
Based on the papers, we should see a gain in the whole chr8 and chr12, not just one arm or specific genes, so we expect a high value for the proportion. 
We see this for chr8 but not in chr12. 
However, when you look at the heatmap you see that most of Chr8 is included in the gain, but only a small part of Chr12 has a gain, so these plots are not surprising.

I think with using the combination of marker gene expression and CNVs, the bottom group of cells represent the tumor cells.

@@

- If we have a known group of normal cells, `InferCNV` can be helpful in identifying cells that could be tumor cells, e.g., cells that have a high number of CNVs. 
- Using `InferCNV` without a group of normal cells does not seem feasible for this particular sample since all tumor cells appear to have the same set of CNVs with little variation. 
- We can define tumor cells by setting a cutoff for the number of CNVs that must be present in a cell, but I'm not sure how transferrable this is to other samples. 
- `InferCNV` can be helpful in looking at specific CNVs that are expected in a tumor type, something that is harder to look at in a tool like `CopyKAT`.
